cardinalR: Generating interesting high-dimensional data structures

### 2025-05-08 paper-cardinalR.Rmd
Added the package list
Added package references
Updated the distributions

$X_1 \sim \mathcal{U}(-2, 2)$
$X_2 = \exp(\pm s_i \cdot X_1) + \epsilon$,

where:
$\epsilon \sim \mathcal{U}(0, 0.1)$ adds local jitter,

@@

$$
X_j \sim \mathcal{N}(0, 0.1^2), \quad j = 3, \dots, p.
$$

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X_1 \sim \mathcal{U}(-2, 8)\$, \$X_2 = 0.5 \cdot X_1 + \epsilon$. The Branch 2 is generated from $X_1 \sim \mathcal{U}(-6, 2)\$, \$X_2 = -0.5 \cdot X_1 + \epsilon$. $\epsilon \sim \mathcal{U}(0, 0.5)$.

@@

$X_2 = s_i \cdot (X_1 - x_{start}) + y_{start} + \epsilon$ where $s_i$ is a sampled slope from a filtered set and $\epsilon \sim \mathcal{U}(0, 0.2)$.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2)\text{ for }j = 3, \dots, p.
$$

@@

Branch 1 is generated from $X_1 \sim \mathcal{U}(0, 1)$, and $X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon$, where $\epsilon \sim \mathcal{U}(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from $X_{1} \sim \mathcal{U}(-1, 0)$, and $X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon$, where $\epsilon \sim \mathcal{U}(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
$$

@@

$$
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5),
$$

@@

$$
x_j \sim \mathcal{N}(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
$$

@@

$$
x_{i1} \sim \mathcal{U}(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim \mathcal{U}(0, 0.5)
$$

@@

Points are sampled independently from $X_1, X_2, X_3 \sim \mathcal{U}(-0.5, 0.5)$ of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in $3\text{-}D$ space.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$

@@

The initial points are drawn from a uniform cube centered at the origin: $X_1, X_2, X_3 \sim \mathcal{U}(-0.5, 0.5)$.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$.

@@

$$
X_i \sim \mathcal{N}_p(\boldsymbol{0}, \Sigma)
$$

@@

$\text{scale}_i \sim \mathcal{U}(-10, 10)$: randomly chosen direction and stretch.

$\text{shift}_i \sim \mathcal{U}(-300, 300)$: large offset to spread features apart.

$\epsilon \sim \mathcal{N}(0, 0.03 \cdot n)$: Gaussian noise scaled to dataset size to introduce mild irregularity.

@@

$$
x_j \sim \mathcal{N}(0, 0.05^2), \quad j = 4, \dots, p.
$$

@@

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let $X_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])$ be the independent variable. A raw polynomial basis of degree 2 is used to form $X_2 = X_1 - X_1^2 + \varepsilon_2$, where $\varepsilon_2 \sim \mathcal{U}(0, 0.5)$.

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let $X_1 \sim \mathcal{U}(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\$, where \$\varepsilon_2 \sim \mathcal{U}(0, 0.5)$.

@@

$$
X_j \sim \mathcal{N}(0, 0.1^2)\$ for \$j = 3, \dots, p.
$$

@@

$$
x_1 \sim \mathcal{U}(-r_x(z),\ r_x(z)), \quad
x_2 \sim \mathcal{U}(-r_y(z),\ r_y(z)), \quad
x_3 \sim \mathcal{U}(-r_x(z),\ r_x(z))
$$.

@@

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1.
$$

@@

$$
u, v \sim \mathcal{U}(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
$$

@@

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
$$

@@

$$
x_p = z \sim \mathcal{U}(0, h).
$$

@@

$$
\theta \sim \text{Uniform sample from 6 hexagon angles}, \quad
r_{\text{point}} \sim \sqrt{\mathcal{U}(0, 1)}.
$$

@@

$$
x_j \sim \mathcal{U}(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right).
$$

@@

$$
T_0 \sim \mathcal{U}(0, 1)^p.
$$

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 4, \dots, p.
$$

@@

In generation of $k$ smaller uniform spheres, each sphere contains $n_2$ points that are sampled uniformly on a sphere with a radius of $r_2$. These spheres are positioned at distinct random locations in $p$-space, with the center of each sphere being drawn from a normal distribution $\mathcal{N}(0, \texttt{loc}^2 I_p)$.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 5, \dots, p.
$$

@@

$$
X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim \mathcal{N}(0, 0.5^2),\text{ and }s_j\text{ is a random scale}.
$$

@@

Let $\theta \sim \mathcal{U}(0, 3\pi)$ represent a random angle on a circular base and $z \sim \mathcal{U}(0, h)$ represent the height along the cylinder.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2),\text{ for }j = 5, \ldots, p.
$$

@@

  * $X_1 = \sin(\phi) \cdot \cos(\theta)$
  * $X_2 = \sin(\phi) \cdot \sin(\theta)$
  * $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim \mathcal{U}(-0.5, 0.5)$ introduces vertical jitter.
  * $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path.

@@

where:
$a_j \in \{2, 3, 4, 5\}$ is a randomly chosen polynomial power,
$s_j \sim \mathcal{U}(0.5, 2)$ is a scale factor,
$\eta_j \sim \mathcal{U}(-\sigma_j, 2\sigma_j)$, with $\sigma_j \sim \mathcal{U}(0, 0.05)$, adds mild randomness.

@@

  * $X_3 = 0.05 \cdot \theta + \varepsilon_3$, with $\varepsilon_3 \sim \mathcal{U}(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix.

@@

$$
X_j \sim \mathcal{N}(0, 0.05^2), \quad j = 5, \dots, p.
$$

@@

The growth pattern resembles a cone, with the height increasing according to $X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim \mathcal{U}(-0.1, 0.6).$

Spiral modulation in the fourth dimension is represented by $X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim \mathcal{U}(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension.

@@

* $X_1 \sim \mathcal{U}(0.1, 2)$: base variable (avoids zero to prevent division errors).
* $X_3 \sim \mathcal{U}(0.1, 0.8)$: independent auxiliary variable.
* $X_2 = \frac{hc}{X_1} + \text{non_fac} \cdot \sin(X_1)$: non-linear combination of **hyperbolic and sinusoidal transformations**, creating sharp curvature and oscillation.
* $X_4 = \cos(\pi \cdot X_1) + \varepsilon$, with $\varepsilon \sim \mathcal{U}(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure.

@@

# Discussion

@@

This article is created using \CRANpkg{knitr} [@yihui2015] and \CRANpkg{rmarkdown} [@yihui2018] in R with the `rjtools::rjournal_article` template.

>>>>

$X_1 \sim U(-2, 2)$
$X_2 = \exp(\pm s_i \cdot X_1) + \epsilon$,

where:
$\epsilon \sim U(0, 0.1)$ adds local jitter,

@@

$$
X_j \sim N(0, 0.1^2), \quad j = 3, \dots, p.
$$

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches 1 and 2 are initialized with fixed slopes and intercepts. The Branch 1 is generated from $X_1 \sim U(-2, 8)\$, \$X_2 = 0.5 \cdot X_1 + \epsilon$. The Branch 2 is generated from $X_1 \sim U(-6, 2)\$, \$X_2 = -0.5 \cdot X_1 + \epsilon$. $\epsilon \sim U(0, 0.5)$.

@@

$X_2 = s_i \cdot (X_1 - x_{start}) + y_{start} + \epsilon$ where $s_i$ is a sampled slope from a filtered set and $\epsilon \sim U(0, 0.2)$.

@@

$$
X_j \sim N(0, 0.05^2)\text{ for }j = 3, \dots, p.
$$

@@

Branch 1 is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1 \cdot X_1 + 1 \cdot X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane.

Branch 2 is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1 \cdot X_{1} - 2 \cdot X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane.

@@

$$
X_j \sim N(0, 0.05^2), \quad \text{for } j = 3, \dots, p.
$$

@@

$$
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5),
$$

@@

$$
x_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}.
$$

@@

$$
x_{i1} \sim U(0, 2), \quad
x_{i2} = -s_i \cdot x_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)
$$

@@

Points are sampled independently from $X_1, X_2, X_3 \sim U(-0.5, 0.5)$ of a cube of side length 1. This results in a uniform cloud of points filling the unit cube in $3\text{-}D$ space.

@@

$$
X_j \sim N(0, 0.05^2), \quad j = 4, \dots, p.
$$

@@

The initial points are drawn from a uniform cube centered at the origin: $X_1, X_2, X_3 \sim U(-0.5, 0.5)$.

@@

$$
X_j \sim N(0, 0.05^2), \quad j = 4, \dots, p.
$$.

@@

$$
X_i \sim N_p(\boldsymbol{0}, \Sigma)
$$

@@

$\text{scale}_i \sim U(-10, 10)$: randomly chosen direction and stretch.

$\text{shift}_i \sim U(-300, 300)$: large offset to spread features apart.

$\epsilon \sim N(0, 0.03 \cdot n)$: Gaussian noise scaled to dataset size to introduce mild irregularity.

@@

$$
x_j \sim N(0, 0.05^2), \quad j = 4, \dots, p.
$$

@@

The curve is constructed by drawing uniformly spaced inputs and applying a second-degree polynomial transformation. Let $X_1 \sim U(\text{range}[1], \text{range}[2])$ be the independent variable. A raw polynomial basis of degree 2 is used to form $X_2 = X_1 - X_1^2 + \varepsilon_2$, where $\varepsilon_2 \sim U(0, 0.5)$.

@@

The shape is generated using a third-degree raw polynomial basis expansion. Let $X_1 \sim U(\text{range}[1], \text{range}[2])$ be the base input. A cubic transformation with vertical jitter is used to define $X_2 = X_1 + X_1^2 - X_1^3 + \varepsilon_2\$, where \$\varepsilon_2 \sim U(0, 0.5)$.

@@

$$
X_j \sim N(0, 0.1^2)\$ for \$j = 3, \dots, p.
$$

@@

$$
x_1 \sim U(-r_x(z),\ r_x(z)), \quad
x_2 \sim U(-r_y(z),\ r_y(z)), \quad
x_3 \sim U(-r_x(z),\ r_x(z))
$$.

@@

$$
x_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1.
$$

@@

$$
u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v.
$$

@@

$$
x_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right).
$$

@@

$$
x_p = z \sim U(0, h).
$$

@@

$$
\theta \sim \text{Uniform sample from 6 hexagon angles}, \quad
r_{\text{point}} \sim \sqrt{U(0, 1)}.
$$

@@

$$
x_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right).
$$

@@

$$
T_0 \sim U(0, 1)^p.
$$

@@

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p.
$$

@@

In generation of $k$ smaller uniform spheres, each sphere contains $n_2$ points that are sampled uniformly on a sphere with a radius of $r_2$. These spheres are positioned at distinct random locations in $p$-space, with the center of each sphere being drawn from a normal distribution $N(0, \texttt{loc}^2 I_p)$.

@@

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p.
$$

@@

$$
X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}.
$$

@@

Let $\theta \sim U(0, 3\pi)$ represent a random angle on a circular base and $z \sim U(0, h)$ represent the height along the cylinder.

@@

$$
X_j \sim N(0, 0.05^2),\text{ for }j = 5, \ldots, p.
$$

@@

  * $X_1 = \sin(\phi) \cdot \cos(\theta)$
  * $X_2 = \sin(\phi) \cdot \sin(\theta)$
  * $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter.
  * $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path.

@@

where:
$a_j \in \{2, 3, 4, 5\}$ is a randomly chosen polynomial power,
$s_j \sim U(0.5, 2)$ is a scale factor,
$\eta_j \sim U(-\sigma_j, 2\sigma_j)$, with $\sigma_j \sim U(0, 0.05)$, adds mild randomness.

@@

  * $X_3 = 0.05 \cdot \theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix.

@@

$$
X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p.
$$

@@

The growth pattern resembles a cone, with the height increasing according to $X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$

Spiral modulation in the fourth dimension is represented by $X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension.

@@

* $X_1 \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors).
* $X_3 \sim U(0.1, 0.8)$: independent auxiliary variable.
* $X_2 = \frac{hc}{X_1} + \text{non_fac} \cdot \sin(X_1)$: non-linear combination of **hyperbolic and sinusoidal transformations**, creating sharp curvature and oscillation.
* $X_4 = \cos(\pi \cdot X_1) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure.

@@

# Conclusion

@@

These `R` packages were used for this work: `cli` [@gabor2025], `tibble` [@kirill2023], gtools [@gregory2023], dplyr [@hadley2023], `stats` [@core2025], `tidyr` [@hadley2024], `purrr` [@hadley2025], `mvtnorm` [@alan2009], `geozoo` [@barret2016], and `MASS` [@venables2002]. This article is created using \CRANpkg{knitr} [@yihui2015] and \CRANpkg{rmarkdown} [@yihui2018] in R with the `rjtools::rjournal_article` template.
